---
title: "Class12_online"
output: pdf_document
date: "2026-02-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Proportion og G/G in a population

Downloaded a CVS file from ENsemble < 
https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39835097-39955098;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel

Here we may read this CVS file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```
```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population. I picked the GBR.
```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G
```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```
This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.
Let's now dig into this further.

## Section 4: Population Scale Analysis [HOMEWORK]
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

How many samples do we have?
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)

```

Lets make a boxplot
```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

G/G expression is much lower than the expression of A/A, therefore, allele A could increase expression. Yes, the SNP does since there is no overlap in the medians.