---
title: "class18_pertussis"
author: Eloise Simpson
format: pdf
editor: visual
---

## Background

Pertussis (a.k.a. whooping cough) is a common lung infection caused by the bacteria B. Pertussis.

This can infect all ages but is most severe for those under 1 year of age.

The CDC track the number of [reported cases in the US](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html) We can scrape this data with the **datapasta** package.

```{r}
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2023L,
    2024L,
    2025L
  ), cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    7063,
    22538,
    21996
  )
       )
```

> Q. Make a plot of `year` vs `cases`

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
  geom_line() 
```

> Q. Add some major milestones including the first wP vaccine roll-out (1946), the switch to the newer aP vaccine (1996), the COVID years (2020).

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col="blue", lty=2) +
  geom_vline(xintercept = 1996, col="pink", lty=2) +
  geom_vline(xintercept = 2020, col="green", lty=2)
```

There were high case numbers in the pre 1940s

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

Cases drop significantly after the implementation of the wP vaccine, but when the aP vaccine is implemented, cases increase.

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

Cases rose in the 1990s because there was a switch to the safer aP vaccine, however, it was less effective in terms of its longevity of protection which is shorter. It can also still be transmitted without strong symptoms by vaccinated individuals and therefore could still circulate. Around 2010, the implementation of boosters brought the cases back down again.

**Why is this vaccine-preventable disease on the upswing?** To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

#CMI-PB project [Computational Models of Immunity - Pertussis boost](https://www.cmi-pb.org) project aims to provide the scientific community with this very information.

They make their data available via JSON format returning API. We can read this in R with the `read_json()` function from the **jsonlite** package:

```{r}
library(jsonlite)
subject <- read_json("http://cmi-pb.org/api/v5_1/subject", simplifyVector=TRUE)

head(subject)
```

> Q4. How many wP and aP individuals are in this `subject` table?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)? Is it representative of the US population?

```{r}
table(subject$race, subject$biological_sex)
```
No this is absolutely not representative of the general US population.

Let's read some more database tables:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)

head(specimen)
```

```{r}
head(ab_titer)
```

To analyse this data we need to first "join" (merge/link) the different tables so we have all the data in one place not spread across different tables.

We can use the `*_join()` family of functions from **dplyr** to do this.

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q. What antibody isotypes are measured for these patients?

```{r}
table(abdata$isotype)
```

> Q. What antigens are reported?

```{r}
table(abdata$antigen)
```

Let's focus on IgG antigen and make a plot of MFI_normalised for all antigens.

```{r}
igg <- abdata |>
  filter(isotype == "IgG")

head(igg)
```

```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

> Q. Is there a difference for aP vs wP individuals with these values?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
  facet_wrap(~infancy_vac)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()

```

> Q. Is there a temporal response - i.e do values increase or decrease over time?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() + 
  facet_wrap(~visit)

```

## Focus on "PT" Pertussis Toxin Antigen

```{r}
pt.igg.21 <- igg |> filter (antigen == "PT",
                     dataset == "2021_dataset")
```

```{r}
ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost, MFI_normalised, col = infancy_vac, group = subject_id) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 14, col = "grey", lty=2)
```

First evidence of significant difference in immune responses between two vaccines.

Add plot from slides, geom_smooth.

```{r}
ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost, MFI_normalised, col = infancy_vac, group = subject_id) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 14, col = "grey", lty=2) +
  geom_smooth(aes(group = infancy_vac, color = infancy_vac), se=FALSE, linewidth = 2) +
  labs(x = "Day Relative to Boost", y = "Normalized MFI", color = "Infancy Vaccine")
```


> Q18. Does this trend look similar for the 2020 dataset?

No, the normalised MFI for the wP vaccine is much lower, even though the aP results are similar.




