---
title: "class17_PCA"
format: pdf
editor: visual
---

## Core Unix commands

Most unix commands have super short names, which makes them quick
to type but annoying to learn. Major file system related
commands include.

pwd: Print working directory
ls: List files and directories
cd: Change Directories
mkdir: Make directories
rm: remove files and directories (delete)
cp: copy files (source > destination)
mv: Move files (basically re-name)
nano: A wee text editor (very basic but always available)

curl:     Download files
wget:     Download files from the web
tar -zxvf:UnTar (unpackage) Tar archive files
gunzip :  UnZip files
$PATH:    The places (dirs) to look for programs

## AWS EC2 Instance

Connect to my instance with:
ssh -i ~/Downloads/bimm143_eloise.pem ubuntu@ec2-54-202-63-98.us-west-2.compute.amazonaws.com

Github link:
https://github.com/EloiseSimpson/Bimm143_githhub/tree/main

Secure copy files between machines, in this case from our instance to our laptop
scp -i ~/Downloads/bimm143_eloise.pem ubuntu@ec2-54-202-63-98.us-west-2.compute.amazonaws.com:/home/ubuntu/work/Bimm143_githhub/Class16/results.txt .

## Class 17 instance
ssh -i "bimm143_eloise.pem" ubuntu@ec2-52-33-145-234.us-west-2.compute.amazonaws.com

export KEY=~/Downloads/bimm143_eloise.pem
export SERVER=ubuntu@ec2-52-33-145-234.us-west-2.compute.amazonaws.com

ssh -i $KEY $SERVER
scp -r 


## Downstream Analysis

```{r}
library(tximport)

base_path <- "class17"
folders <- dir(path = base_path, pattern = "SRR21568*", full.names = TRUE)
samples <- sub("_quant$", "", basename(folders))
files <- file.path(folders, "abundance.h5")
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```
```{r}
head(txi.kallisto$counts)
```
```{r}
colSums(txi.kallisto$counts)
```
```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## Principal Component Analysis

```{r}
pca <- prcomp(t(x), scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```
```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()
```

